---
title: "Results manuscript"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# General stats

Total number of reads: `r general_stats$N_reads`

Total number of OTUs: `r general_stats$N_OTU`

Number of OTUs in Bottom: `r general_stats$N_OTU_BOT`

Number of OTUs in Top: `r general_stats$N_OTU_TOP`

Number of OTUs intersection Top-Bottom: `r general_stats$N_OTU_TOP_INTER_BOT`

Number of OTUs in Bottom only: `r general_stats$N_OTU_BOT_ONLY`, `r ((general_stats$N_OTU_BOT_ONLY/general_stats$N_OTU)*100) %>% round(2)`%

Number of OTUs in Top only: `r general_stats$N_OTU_TOP_ONLY`, `r ((general_stats$N_OTU_TOP_ONLY/general_stats$N_OTU)*100) %>% round(2)`%

Number OTUs classified taxonomy: `r general_stats$N_OTU - general_stats$N_OTU_UNCLASSIF_TAX`, `r (((general_stats$N_OTU - general_stats$N_OTU_UNCLASSIF_TAX) / general_stats$N_OTU)*100) %>% round(2)`%

Number OTUs classified trophic group: `r general_stats$N_OTU - general_stats$N_OTU_UNCLASSIF_TROPH`, `r (((general_stats$N_OTU - general_stats$N_OTU_UNCLASSIF_TROPH) / general_stats$N_OTU)*100) %>% round(2)`%


# Rarified Richness

### OTUs

```{r}
rar_richness %>% 
  filter(INVENTORY == "OTU") %>% 
  .$MEAN_RAR_RICHNESS
rar_richness %>% 
  filter(INVENTORY == "OTU") %>% 
  .$TEST_RAR_RICHNESS
```

### Taxonomic group (rank 2)
```{r}
rar_richness %>% 
  filter(INVENTORY == "Taxonomie Rank 2") %>% 
  .$MEAN_RAR_RICHNESS
rar_richness %>% 
  filter(INVENTORY == "Taxonomie Rank 2") %>% 
  .$TEST_RAR_RICHNESS
```



# NMDS Plot
```{r, out.width="1000px", dpi=120}
FIG_nmds_main <- nmds_rar_list %>% 
  filter(INVENTORY == "OTU") %>% 
  .$SITE_SCORES %>% 
  .[[1]] %>% 
  left_join(lakes_meta_nr, by = "LAC_POS") %>% 
  left_join(lakes_bv, by = "LAC") %>% 
  filter(!is.na(LAC)) %>% 
  mutate(LABS_LAC_BOTTOM = ifelse(POSITION == "Bottom", LAC_CODE, "")) %>% 
  ggplot() +
  geom_line(aes(x = NMDS1, y = NMDS2, group = LAC), color = "lightgrey") +
  #ggforce::geom_mark_ellipse(aes(x = NMDS1, y = NMDS2, fill = POSITION), ) +
  stat_ellipse(aes(x = NMDS1, y = NMDS2, fill = POSITION), geom = "polygon", level = 0.95, alpha = 0.2) +
  geom_point(aes(x = NMDS1, y = NMDS2, shape = POSITION, color = POSITION), size = 2) +
  geom_text(aes(x = NMDS1, y = NMDS2, label = LABS_LAC_BOTTOM), size = 2, nudge_y = -0.04) +
  labs(title = "") +
  theme_bw() +
  theme(axis.text.y.right = element_text(size = 6, angle = 45),
        axis.text.x.top = element_text(size = 6, angle = 45, hjust = 0, vjust = 0),
        legend.position = "none")

FIG_nmds_bray <- TB_pw_BC_plots %>% 
  filter(INVENTORY == "OTU") %>% .$hist_plots %>% .[[1]]

FIG_nmds_intra <- beta_div_TB %>% 
  filter(INVENTORY == "OTU") %>% .$TB_intra_plots %>% .[[1]]


# ggsave("figures/print/FIG_nmds_main.pdf", FIG_nmds_main, device = "pdf", width = 13.1, height = 10, units = "cm", scale = 1)
# ggsave("figures/print/FIG_nmds_bray.pdf", FIG_nmds_bray, device = "pdf", width = 4.7, height = 4.7, units = "cm", scale = 1)
# ggsave("figures/print/FIG_nmds_intra.pdf", FIG_nmds_intra, device = "pdf", width = 4.7, height = 4.7, units = "cm", scale = 1)

FIG_nmds_main
FIG_nmds_bray
FIG_nmds_intra
```

NMDS Stress: `r nmds_rar_list %>% filter(INVENTORY == "OTU") %>% .$NMDS %>% .[[1]] %>% .$stress`

# PERMANOVA (change in location)
```{r}
dat <- com_rar_list %>% 
  filter(INVENTORY == "OTU") %>% .$data %>%
  .[[1]] %>% 
  spread_cdm(LAC_POS, TAXON, COUNT)

cond <- ifelse(str_detect(rownames(dat), "_Bottom$"), "Bottom", "Top")
blocks <- gsub("_Top|_Bottom", "", rownames(dat))

adonis_blocks(dat, cond, permutation = 1, method = "bray", blocks = blocks)

```


# Homogeneity of multivariate dispersions (change in dispersion)
```{r, fig.height = 4, fig.width = 5}
beta_div_TB %>% 
  filter(INVENTORY == "OTU") %>% .$BETA_DISP %>% .[[1]]
beta_div_TB %>% 
  filter(INVENTORY == "OTU") %>% .$BETA_DISP_TEST %>% .[[1]]

```

# Effect of altitude on the magnitude of community changes

### Decision tree model
```{r}
dat <- com_rar_list %>% 
  filter(INVENTORY == "OTU") %>% .$BRAY_DIST %>%
  .[[1]] %>% 
  filter(LAC_POS_1 != LAC_POS_2, LAC_1 == LAC_2) %>% 
  left_join(select(lakes_bv, LAC, ALTITUDE),
            by = c("LAC_1" = "LAC"))

tr_mod_alt <- rpart(BRAY_DIST ~ ALTITUDE, data = dat, model = TRUE)

tr_mod_alt

alt_split <- tr_mod_alt$splits[1, "index"]

bray_split <- dat %>%
  mutate(ALT_SPLIT = ifelse(ALTITUDE > alt_split, "High altitude", "Low altitude")) %>% 
  group_by(ALT_SPLIT) %>% 
  summarise(mean_BRAY = mean(BRAY_DIST),
            sd_BRAY = sd(BRAY_DIST),
            n_BRAY = n()) %>% 
  mutate(CI_BRAY_SUP = mean_BRAY + 1.96 * sd_BRAY / sqrt(n_BRAY),
         CI_BRAY_INF = mean_BRAY - 1.96 * sd_BRAY / sqrt(n_BRAY))

wilcox.test(dat$BRAY_DIST[dat$ALTITUDE > alt_split], dat$BRAY_DIST[dat$ALTITUDE < alt_split])
bartlett.test(dat$BRAY_DIST, ifelse(dat$ALTITUDE > alt_split, "High altitude", "Low altitude"))
var.test(dat$BRAY_DIST[dat$ALTITUDE < alt_split], dat$BRAY_DIST[dat$ALTITUDE > alt_split])

FIG_alt_tree_regression <- dat %>%
  ggplot() +
  geom_rect(aes(xmin = 0, xmax = alt_split,
                ymin = bray_split$CI_BRAY_INF[2], ymax = bray_split$CI_BRAY_SUP[2]), fill = "grey80") +
  geom_rect(aes(xmin = alt_split, xmax = max(dat$ALTITUDE),
                ymin = bray_split$CI_BRAY_INF[1], ymax = bray_split$CI_BRAY_SUP[1]), fill = "grey80") +
  geom_point(aes(ALTITUDE, BRAY_DIST)) +
  geom_segment(aes(x = 0, y = bray_split[2, "mean_BRAY", drop = TRUE],
                   xend = alt_split, yend = bray_split[2, "mean_BRAY", drop = TRUE]), size = 1) +
  geom_segment(aes(x = alt_split, y = bray_split[1, "mean_BRAY", drop = TRUE],
                   xend = max(dat$ALTITUDE), yend = bray_split[1, "mean_BRAY", drop = TRUE]), size = 1) +
  xlab("Altitude (m)") + ylab("Past-Recent Bray-Curtis dissimilarity") +
  theme_bw()

# ggsave("figures/print/FIG_alt_tree_regression.pdf", FIG_alt_tree_regression,
#        device = "pdf", width = 8.7, height = 8.7, units = "cm", scale = 1)

FIG_alt_tree_regression

```



# Differential abundance analyses
# ANOVA
```{r}
deseq_TB_AOV1 %>% 
  filter(INVENTORY == "Taxonomie Rank 1") %>% .$barplots %>% .[[1]]

FIG_deseq_taxo2 <- deseq_TB_AOV1 %>% 
  filter(INVENTORY == "Taxonomie Rank 2") %>% .$barplots %>% .[[1]]

deseq_TB_AOV1 %>% 
  filter(INVENTORY == "Taxonomie Rank 2") %>% .$results %>% .[[1]] %>%
  knitr::kable()


FIG_deseq_troph <- deseq_TB_AOV1 %>% 
  filter(INVENTORY == "Trophic Type") %>% .$barplots %>% .[[1]]

deseq_TB_AOV1 %>% 
  filter(INVENTORY == "Trophic Type") %>% .$results %>% .[[1]] %>%
  knitr::kable()

# ggsave("figures/print/FIG_deseq_taxo2.pdf", FIG_deseq_taxo2,
#        device = "pdf", width = 8.7, height = 13, units = "cm", scale = 1)
# ggsave("figures/print/FIG_deseq_troph.pdf", FIG_deseq_troph,
#        device = "pdf", width = 8.7, height = 4.5, units = "cm", scale = 1)

FIG_deseq_taxo2
FIG_deseq_troph
```


# Sediment Pigments
```{r, fig.height = 5, fig.width = 5}

sed_basics_plots$combined_plots <-
  cowplot::plot_grid(sed_basics_plots$Corg + theme(panel.grid = element_blank()),
                   sed_basics_plots$PIG + theme(panel.grid = element_blank()),
                   sed_basics_plots$DNA + theme(panel.grid = element_blank()),
                   labels = c("A", "B", "C"), ncol = 3, nrow = 1)

 
# ggsave("figures/print/FIG_sediments.pdf", sed_basics_plots$combined_plots,
#        device = "pdf", width = 8.7, height = 8.7, units = "cm", scale = 1)


sed_basics_plots$combined_plots

sed_basics_tests$Corg
sed_basics_tests$PIG
sed_basics_tests$DNA
```

