---
title: "Supplementary Information"
output:
  pdf_document:
    fig_caption: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(magrittr)
load("../data/Supp_mat.Rdata")
fig_num <- 0
tab_num <- 0
ins_tab <- function(){
  .GlobalEnv$tab_num <- .GlobalEnv$tab_num + 1
  paste0("Table S", .GlobalEnv$tab_num, ".")
}
ins_fig <- function(){
  .GlobalEnv$fig_num <- .GlobalEnv$fig_num + 1
  paste0("Figure S", .GlobalEnv$fig_num, ".")
}
library(kableExtra)
usepackage_latex("booktabs")
usepackage_latex("longtable")
usepackage_latex("array")
usepackage_latex("multirow")
usepackage_latex("xcolor", "table")
usepackage_latex("wrapfig")
usepackage_latex("float")
usepackage_latex("colortbl")
usepackage_latex("pdflscape")
usepackage_latex("tabu")
usepackage_latex("threeparttable")
usepackage_latex("threeparttablex")
usepackage_latex("ulem", "normalem")
usepackage_latex("makecell")

```


## Supplementary Information Text
Heading
Subhead. Type or paste text here. You may break this section up into subheads as needed (e.g., one section on “Materials” and one on “Methods”).



<!-- \pagebreak -->
<!-- ```{r, message = FALSE} -->
<!-- lakes_tab %>%  -->
<!--   ggplot(aes(x = Altitude, fill = `Trophic status`)) + -->
<!--     geom_histogram(position = 'stack', binwidth = 500) + -->
<!--     scale_fill_manual(values = c("Eutrophic" = "#006D2C", -->
<!--                                  "Meso-eutrophic" = "#2CA25F", -->
<!--                                  "Mesotrophic" = "#66C2A4", -->
<!--                                  "Oligo-mesotrophic" = "#99D8C9", -->
<!--                                  "Oligotrophic" = "#CCECE6", -->
<!--                                  "Unknown" = "grey70")) + -->
<!--   labs(fill = "Trophic status") + xlab("Altitude (m)") + ylab("Count") + -->
<!--   theme_bw() -->


<!-- ``` -->


<!-- SKIPED NUM Distribution of lake elevation and trophic status. -->


\pagebreak

![](../figures/supplementary_info/SI_FIG_bioinfo.pdf)

**`r ins_fig()`** Tracking of the number of DNA reads and OTUs during the filtering and data pre-processing steps.


\pagebreak
```{r, fig.height = 8}
tax_rich_dat <-
  com_merged_raw_list %>% 
  filter(INVENTORY == "OTU") %>% 
  .$data %>% .[[1]] %>%
  spread_cdm(LAC_POS, TAXON, COUNT) %>% 
  rrarefy(sample = min(rowSums(.))) %>% 
  tidy_cdm(row.name = "LAC_POS") %>% 
  left_join(lakes_meta_nr)

tax_rich_lakes <- tax_rich_dat %>% 
  group_by(LAC, POSITION) %>% 
  summarise(RICH_LAKE = sum(COUNT > 0))

p <- tax_rich_dat %>% 
  left_join(otu_meta, by = c("TAXON" = "OTU_ID")) %>% 
  group_by(ADL_RANK_2, POSITION) %>% 
  nest() %>% 
  mutate(RICH = map(data, function(data){
    data %>% 
      spread_cdm(LAC, TAXON, COUNT) %>% 
      specnumber() %>% 
      enframe("LAC", "RICHNESS")
  })) %>% 
  unnest(RICH) %>% 
  left_join(tax_rich_lakes) %>% 
  mutate(REL_RICHNESS = RICHNESS/RICH_LAKE) %>% 
  group_by(ADL_RANK_2, POSITION) %>% 
  summarise(REL_RICHNESS_MEAN = mean(REL_RICHNESS),
            REL_RICHNESS_SD = sd(REL_RICHNESS)) %>% 
  left_join(read_csv("../data/Ranks_Adl.csv"), by = c("ADL_RANK_2" = "RANK2")) %>% 
  group_by(RANK1) %>% 
  mutate(GRP2_REL_RICHNESS_SUM = sum(REL_RICHNESS_MEAN, na.rm = TRUE)) %>%
  ungroup() %>% 
  mutate(ADL_RANK_2 = fct_explicit_na(ADL_RANK_2, "Unclassified"),
         RANK1 = fct_explicit_na(RANK1, "Unclassified")) %>% 
  ggplot() +
  geom_col(aes(fct_reorder(ADL_RANK_2, REL_RICHNESS_MEAN),
               REL_RICHNESS_MEAN, fill = POSITION),
           position = "dodge2") +
  facet_wrap(~ fct_reorder(RANK1, GRP2_REL_RICHNESS_SUM, .desc = TRUE),
             scales = "free_y", ncol = 1, strip.position = "left") +
  geom_linerange(aes(x = fct_reorder(ADL_RANK_2, REL_RICHNESS_MEAN),
                         ymin = REL_RICHNESS_MEAN + REL_RICHNESS_SD,
                         ymax = REL_RICHNESS_MEAN - REL_RICHNESS_SD),
           position = position_dodge2(width = 1, padding = 0.5)) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_discrete(name = "Stratum", labels = c("Past (bottom)", "Recent (top)")) +
  xlab("") + ylab("Relative richness") +
  theme(strip.text.y = element_text(angle = 180),
        strip.placement = "outside")

gp <- ggplotGrob(p)
f_row <- gp$layout$t[grepl("panel", gp$layout$name)]
x_var <- sapply(ggplot_build(p)$layout$panel_scales_x,
                function(l) length(l$range$range))
gp$heights[f_row] <- gp$heights[f_row] * x_var
grid::grid.draw(gp)
```

**`r ins_fig()`**  Mean relative richness (number of distinct OTUs) for each taxa (2^nd^ rank of the classification of Adl et al. 2019) computed for the bottom and the top rarefied samples across all lakes. Horizontal lines show the standard error. The superior rank of each taxa is given in the grey margin for guidance.




<!-- \pagebreak -->
<!-- ```{r, fig.height = 7} -->

<!-- prow <- plot_grid(frac_taxa_temp$plots[[1]] + -->
<!--             theme(legend.position = "none"), -->
<!--           frac_taxa_temp$plots[[2]] + -->
<!--             theme(legend.position = "none") + -->
<!--             xlab(""), -->
<!--           labels = "AUTO") -->
<!-- legend <- get_legend( -->
<!--   frac_taxa_temp$plots[[1]] + -->
<!--     guides(color = guide_legend(nrow = 1)) + -->
<!--     theme(legend.position = "bottom", -->
<!--           legend.title = element_blank()) -->
<!-- ) -->
<!-- plot_grid(prow, legend, ncol = 1, rel_heights = c(1, .1)) -->
<!-- ``` -->

<!-- **`r `**  Proportion of OTUs (*A*) and taxonomic groups (*B*) specific to the bottom (Bottom only) and top (Top only) strata, or common to both strata (Intersection). The total number of OTUs or taxa is given for each lake in parenthesis. -->



\pagebreak
```{r, fig.height = 6}

SUPP_MAT$JACC$FIG_COMB

```


**`r ins_fig()`** Community analyses reproduced with the Jaccard dissimilarity index. (*A*) NMDS of community compositions of the ‘top’ (blue triangles) and ‘bottom’ (red circles) samples with 95% confidence ellipses represented for each group. Grey lines connect ‘top’ and ‘bottom’ samples taken within the same lake. (*B*) Distribution of Jaccard index values computed for each lake between ‘top’ and ‘bottom’ samples . (*C*) Distribution of the distances between samples and group geometric median for ‘top’ (blue) and ‘bottom’ (red) samples. (*D*) Relationship between lake altitude and community dissimilarity (Jaccard index) between the ‘top’ and ‘bottom’ strata. Fitted regression tree model is represented by black lines (mean values). Grey shading represent the 95% confidence intervals around means.


\pagebreak
```{r, fig.height = 6}

SUPP_MAT$TAXO$FIG_COMB

```

**`r ins_fig()`** Community analyses reproduced with the taxonomy table. (*A*) NMDS of community compositions of the ‘top’ (blue triangles) and ‘bottom’ (red circles) samples with 95% confidence ellipses represented for each group. Grey lines connect ‘top’ and ‘bottom’ samples taken within the same lake. (*B*) Distribution of Bray-Curtis index values computed for each lake between ‘top’ and ‘bottom’ samples . (*C*) Distribution of the distances between samples and group geometric median for ‘top’ (blue) and ‘bottom’ (red) samples. (*D*) Relationship between lake altitude and community dissimilarity (Bray-Curtis index) between the ‘top’ and ‘bottom’ strata. Fitted regression tree model is represented by black lines (mean values). Grey shading represent the 95% confidence intervals around means.

\pagebreak
```{r, fig.height = 8}
deseq_TB_AOV1 %>%
  filter(INVENTORY == "Taxonomie Rank 3") %>%
  .$barplots %>% .[[1]] +
  theme(plot.title = element_blank())
```

**`r ins_fig()`** Magnitude of change (expressed in log2 fold change) in data aggregated at the 3^rd^ taxonomy rank of Adl et al. (2019) between the past and recent strata, as estimated by the DESeq2 analysis (Love et al. 2014). Dark bars represent groups for which the change was found significant (corrected p-value < 0.05). Horizontal lines show the standard error.



\pagebreak
```{r, fig.height = 8}
t <- otu_tbl_replicates %>% 
  filter(!LAC %in% c("Léman", "Bourget")) %>% 
  mutate(ID_SAMPLE = str_c(LAC_CODE, POSITION, REPLICATE, sep = "_")) %>% 
  spread_cdm(ID_SAMPLE, OTU_ID, OTU_COUNT) %>% 
  vegdist(method = "bray") %>% 
  hclust() %>% 
  ape::as.phylo.hclust()
t2 <- phytools::splitTree(t, list(node = 188, bp = 0))
par(mfrow = c(1, 2))
ape::plot.phylo(ape::drop.tip(t, t2[[1]]$tip.label), cex = 0.5, no.margin = TRUE)
ape::plot.phylo(ape::drop.tip(t, t2[[2]]$tip.label), cex = 0.5, no.margin = TRUE)
par(mfrow = c(1, 1))
```

**`r ins_fig()`** Hierarchical clustering of the OTU matrix (Bray Curtis dissimilarity) including the technical replicates. The code of each sample corresponds to the combination of the lake (three-letter code; see Supplementary Table S1), stratum (Top or Bottom) and replicate number (1 or 2). Lakes Bourget and Léman are not represented as they have only one replicate. For better readability, the tree has been divided in two parts.



<!-- TABLE SECTION STARTS HERE -->

\pagebreak
```{r, message = FALSE}
lakes_kml <- lakes_kml[lakes_kml$Name %in% otu_merge_tbl$LAC, ]

lakes_tab <- read_csv("../data/lakes_coring.csv", na = character()) %>% 
  set_names(c("Code", "Name", "Longitude", "Latitude", "Altitude", "Depth\n(max)",
              "Trophic status", "Coring\nyear", "XrF", "$^{14}$C", "Radionuclides", "included_DNA")) %>% 
  mutate(Radionuclides = str_replace(Radionuclides, "137Cs", "$^{137}$Cs"),
         Radionuclides = str_replace(Radionuclides, "210Pb", "$^{210}$Pb")) %>% 
  arrange(included_DNA, Altitude)

not_included_min <- min(which(lakes_tab$included_DNA == 2))
not_included_max <- max(which(lakes_tab$included_DNA == 2))

lakes_tab <- select(lakes_tab, -included_DNA)

kableExtra::kable(lakes_tab, "latex", digits = 3, booktabs = TRUE, linesep = "", escape = FALSE,
                  col.names = linebreak(colnames(lakes_tab), align = "c")) %>% 
  kable_styling(latex_options = c("scale_down")) %>% 
  pack_rows("Not included for DNA analyses", not_included_min, not_included_max, indent = FALSE, latex_gap_space = "0.5em")

rm(not_included_min, not_included_max)

```

**`r ins_tab()`** List of the lakes included in the study with their three-letter code, geographic coordinates (decimal degrees; WGS84), altitude (meters) and maximum depth (meters). ADD TROPHIC !!!!!!!


\pagebreak
```{r, fig.height = 6}
deseq_TB_AOV1 %>% 
  filter(INVENTORY == "Trophic Type") %>% .$results %>% .[[1]] %>%
  mutate(Taxa = str_replace(Taxa, "<NA>", "Unclassified (NA)")) %>%
  select(-lfcSE, -pvalue) %>% 
  knitr::kable("latex", digits = 3, booktabs = TRUE, linesep = "", escape = FALSE,
               col.names = linebreak(c("Trophic Group", "Base Mean",
                                       "log2 Fold Change", "Stat", "Padj"), align = "c")) %>% 
  kable_styling(position = "center")
```


**`r ins_tab()`** Detailed results of the DESeq2 analysis conducted on data aggregated by trophic groups. For each group, column *Base Mean* indicates the average of the normalized count values, dividing by size factors, *log2 Fold Change* is the magnitude of change between the past and recent strata, column *stat* is the Wald statistic and *padj* the p-value of the test corrected with the Benjamini & Hochberg method.



\pagebreak
```{r, fig.height = 6}
deseq_TB_AOV1 %>% 
  filter(INVENTORY == "Taxonomie Rank 3") %>% .$results %>% .[[1]] %>%
  left_join(select(otu_taxo_adl_conv, ADL_RANK_1, ADL_RANK_2, ADL_RANK_3), by = c("Taxa" = "ADL_RANK_3")) %>% 
  filter(!duplicated(Taxa)) %>% 
  select(Rank_1 = ADL_RANK_1, Rank_2 = ADL_RANK_2, Rank_3 = Taxa, baseMean, log2FoldChange, stat, padj) %>% 
  arrange(Rank_1, Rank_2, Rank_3) %>% 
  mutate(Rank_1 = ifelse(is.na(Rank_1), "Unclassified", Rank_1),
         Rank_2 = ifelse(is.na(Rank_2), "Unclassified", Rank_2)) %>% 
  mutate(Rank_1 = ifelse(!duplicated(Rank_1), Rank_1, ""),
         Rank_2 = ifelse(!duplicated(Rank_2), Rank_2, "")) %>% 
  knitr::kable("latex", digits = 3, booktabs = TRUE, linesep = "", escape = FALSE, longtable = TRUE,
               col.names = linebreak(c("Rank 1", "Rank 2", "Rank 3", "Base Mean",
                                       "log2 Fold\nChange", "Stat", "Padj"), align = "c")) %>% 
  add_header_above(c("Taxonomic rank (Adl et al. 2019)" = 3, " " = 4)) %>% 
  kable_styling(latex_options =c("repeat_header"))

```


**`r ins_tab()`** Detailed results of the DESeq2 analysis conducted on data aggregated at the 3^rd^ taxonomy rank of Adl et al. (2019). Ranks 1 and 2 are given for guidance. For each taxa, column *Base Mean* indicates the average of the normalized count values, dividing by size factors, *log2 Fold Change* is the magnitude of change between the past and recent strata, column *stat* is the Wald statistic and *padj* the p-value of the test corrected with the Benjamini & Hochberg method.





\pagebreak

**Dataset S1** Complete list of OTUs used in the analyses with their DNA sequence, their taxonomic classification and their trophic classification.


## References


